Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2016-09-05 pkg/glmertree/paper/treatment_interactions.Rnw
Small touch-ups in paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@963 255f4249-2263-4417-b1ea-8498779e27a7

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but datasets suited for their detection may often have a clustered or multilevel structure. For example, observations in a dataset may be nested within research centers or trials. Therefore, we propose generalized linear mixed-effects model trees (GLMM tree), allowing for the detection of treatment-subgroup interactions, as well as taking into account the nested structure of a dataset. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with thattrees without random effects, and GLMMs with pre-specified interactions. GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of datasets, and trees without random effects in 61\% of datasets. GLMM tree also outperformed trees without random effects in terms of predictive accuracy (.94 and .88 on average, respectively) and Type-I error rates (4 and 33\%, respectively). Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal accuracy in predicting treatment-effect differences, but better accuracy in detecting interactions. We illustrate the application of GLMM tree on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several tree-based algorithms to the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, which allows users to select from a range of statistical methods, including trees. As such, the \citeA{ZhanyTsia12Stat} method can also be referred to as a tree-based method.

Often, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. In tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection, also take into account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM tree.

@@

In what follows, we introduce an artificial motivating dataset, which will be used to illustrate and explain treatment-subgroup interaction detection with GLM-based MOB and GLMM tree. In the Empirical Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects. In the Application, we apply the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. Finally, in the Discussion we summarize the results and point out some directions for future research.

@@

For example, let us take a global GLM to estimate the overall treatment effect in the artificial motivating dataset. The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function\footnote{An overview of notation used is provided in the appendix.}:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment group. Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper; that is, two treatment conditions and no additional covariates are included in the GLM. Furthermore, we assume a continuous response variable, employing a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$. However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.

@@

The MOB algorithm detects such moderation, by testing for parameter stability over a set of auxiliary covariates or partitioning variables. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function. By definition, the empirical scores of all observations in a dataset sum to zero. When the model is correctly specified, the expected value of the score for each observation is also zero. Therefore, under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$ \cite<c.f., >{MerkyZeil13}. To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.

@@

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on a GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node. As a result, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':
%
\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}
%
Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all coefficients interact with the subgroup indicator. Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{*}$ would have length $2J$, and contain the subgroup-specific fixed-effects coefficients.

@@

Figure~\ref{fig:example_mobtree} shows the GLM tree grown on the artificial motivating dataset. By using the three auxiliary covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the treatment-subgroup interactions as described above. However, the two subgroups in the middle do not represent true subgroups, and this may be due to the clustered structure of the dataset not being taken into account.

@@

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a GLMM would be more appropriate. The model in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

For simplicity, we assume that only cluster-specific intercepts are included in the model in the current paper. However, multiple random-effects coefficients can easily be included.

@@

Although GLMMs are adequately suited for taking into account the nested structure of a datasets, the global fixed-effects model in Equation \ref{eq:mixedeffects} may not describe the data well. Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation \ref{eq:fixedeffects_MOB} and random effects are incorporated as well: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random effects.

@@

R \cite{R14} was used for generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions. For other response distributions, the \verb|glmtree| function would be available. For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.

@@

\subsection{Treatment-subgroup interaction design} 
For generating datasets with treatment-subgroup interactions, we used one of the treatment-subgroup interaction designs from \citeA{DussyMech14}, which is depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by values on the (true) partitioning variables $U_2$, and $U_1$ or $U_5$. The other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

@@

\begin{enumerate} 
\item Three levels for sample size: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning variable (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of $\approx 0.42$), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of $\approx 0.42$)\footnote{Note that, when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition. However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.}. 
\item Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes. In datasets with treatment-subgroup interactions, the treatment effect difference varied only in nodes 4 and 7, with levels $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$). 
\end{enumerate}

\subsection{Variable distributions}
Following the approach of \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$. The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ have the same standard deviation: ${\sigma_U}_k = 10$. Correlations between $U$ variables vary according to the third facet of the data-generating design described above.

@@

\subsection{Assessment of performance}

@@

Figure~\ref{fig:xyplot_treesize_interact} indicates that average tree size increases slightly with sample size, which may be due to increased power. In the absence of random effects (i.e., $\sigma_{b}=0$), both GLM and GLMM trees have about 7 nodes on average. However, clear differences in tree size were observed when $\sigma_b > 0$. When sample size is small and random intercepts are not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average. When random intercepts are not correlated with a partitioning covariate and sample size is large (i.e., 500 or 1000), GLM and GLMM trees are about equally sized and have approximately the true tree size, on average. When random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10). This effect was somewhat stronger when the random intercept was correlated with a covariate that did not appear in the true tree.

@@

To assess the accuracy of trees recovered by both algorithms, we inspected variables and values that were selected for partitioning. For the first split, GLMM tree selected the true partitioning variable ($U_2$) in all datasets, and GLM tree in all but one datasets. The splitting value for $U_2$ selected for the first split was 29.94 for both GLM and GLMM tree, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}). GLM tree did show somewhat higher variability in recovering the splitting value for the first split than GLMM tree ($\mathrm{SD}=0.154$ and $\mathrm{SD}=0.127$, respectively).

GLMM tree accurately recovered the partition in 90.40\% of datasets, whereas GLM tree accurately recovered the partition in only 61.44\% of datasets. To find the most important predictors of tree accuracy, we fitted a GLM with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors of tree accuracy were algorithm type, sample size, variance of the random intercept, and the level of dependence between the partitioning variables and the random intercept. These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_treeaccuracy}).

Figure~\ref{fig:xyplot_treeaccuracy} indicates that in the absence of random effects, the trees recovered by GLM and GLMM tree were about equally accurate. In the presence of random effects, GLM trees were much less accurate than GLMM trees. This was found for all sample sizes, when random effects were correlated with a partitioning covariate. This effect was somewhat stronger when the correlated $U_k$ was not a true partitioning variable. When random intercepts were not correlated with one of the $U_k$ variables, GLMM tree clearly outperformed GLM tree only when sample size was small (i.e., $N = 200$).

@@

\subsubsection{Predictive accuracy of trees}
To assess predictive accuracy, correlations between the true and predicted treatment-effect differences of both algorithms were calculated in test datasets. The predicted treatment-effect differences of GLMM tree were closer to the true differences than those of GLM tree, with a mean correlation of .93 ($\mathrm{SD}=0.12$) for GLMM tree and .88 ($\mathrm{SD}=0.19$) for GLM tree. An ANOVA indicated that the most important predictors of predictive accuracy were algorithm type, sample size, variance of the random intercept, and the treatment difference effect size. These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_correlations}).

Figure~\ref{fig:xyplot_correlations} shows clear main effects of sample size and treatment-effect differences for both algorithms. In the absence of random effects (i.e., $\sigma_b = 0$), predictions of GLM and GLMM tree were about equally accurate. However, it should be noted that in the absence of random effects, GLMM tree tends to outperform GLM tree in smaller sample sizes (i.e., 200), whereas GLM tree tends to outperform GLMM tree in larger sample sizes (i.e., 500 and 1000). In the presence of random effects, GLM tree predictions were much less accurate than those of GLMM tree when treatment-effect differences were smaller (i.e., Cohen's $d = .5$). This effect was also observed, but much less pronounced, when treatment-effect differences were larger (i.e., Cohen's $d = 1.0$). In addition, with increasing sample size and treatment-effect differences, the difference in predictive accuracy between GLM and GLMM tree becomes smaller.

@@

\subsubsection{Design}
For assessing Type-I error of GLM and GLMM tree, we generated datasets without treatment-subgroup interactions, in which there is only a main effect of treatment in the population. Put differently, the number of subgroups or terminal nodes in these datasets was $J=1$, and there was only a single value of $\beta_j = \beta$ in every dataset. For the datasets without treatment-subgroup interactions, we varied the same parameters of the data-generating process as in Study I. However, the sixth facet of the data-generating process had only two levels for these datasets, as there were no `true' partitioning variables in these datasets. Therefore, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2$ = 21,600 datasets without treatment-subgroup interactions were generated.

>>>>

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available tree-based algorithms allow for taking into account a clustered or nested structure. Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, as well accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of trees without random effects, and GLMMs with pre-specified interactions. GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of datasets, and trees without random effects in 61\% of datasets. GLMM tree also outperformed trees without random effects in terms of predictive accuracy (.94 and .88 on average, respectively) and Type-I error rates (4 and 33\%, respectively). Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal predictive accuracy (.60 on avergae), but better accuracy in detecting interactions. We illustrate the application of GLMM tree on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are particularly useful for exploratory purposes, as they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions. Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.

In many instances, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. For tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection take into account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM tree.

@@

In what follows, we first introduce an artificial motivating dataset, which will be used to illustrate and explain treatment-subgroup interaction detection with GLM-based MOB and GLMM tree. In the Empirical Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects. In the Application, we apply the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. Finally, in the Discussion we summarize the results and point out some directions for future research.

@@

For example, let us take a global GLM to estimate the overall treatment effect in the artificial motivating dataset. The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function\footnote{An overview of notation used is provided in the appendix.}:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups. Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper. Also, we assume a continuous response variable, employing a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$. However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.

@@

The MOB algorithm can be used to detect such moderation, by testing for parameter stability over a set of auxiliary covariates or partitioning variables. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function. When the model is correctly specified, the expected value of the score for each observation equals zero. Therefore, under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when observations are ordered by the values of a potential partitioning variable $U_k$ \cite<c.f., >{MerkyZeil13}. To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.

@@

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node or subgroup:
%
\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}
%
Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all coefficients interact with the subgroup indicator. Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{*}$ would have length $2J$, and contain the subgroup-specific fixed-effects coefficients.

@@

Figure~\ref{fig:example_mobtree} shows the GLM tree grown on the artificial motivating dataset. By using the three auxiliary covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the true treatment-subgroup interactions described above. However, the two subgroups in the middle do not represent true subgroups, which may be due to the clustered structure of the dataset not being taken into account.

@@

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a GLMM would be more appropriate. The model in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, each $m$th element representing the random intercept for cluster $m$; again, for simplicity we assume that only cluster-specific intercepts are included in the model, but multiple random-effects coefficients can easily be included. Furthermore, within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

@@

Although the random-effects part of the GLMM in Equation \ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part may not describe the data well. Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation \ref{eq:fixedeffects_MOB} and random effects are incorporated as well: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random effects.

@@

R \cite{R14} was used for generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions. For other response distributions, the \verb|glmtree| function would be available. For estimation of GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.

@@

\subsection{Simulation design}

\subsubsection{Treatment-subgroup interactions} 
For generating datasets with treatment-subgroup interactions, we used one of the treatment-subgroup interaction designs from \citeA{DussyMech14}, which is depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by values on the (true) partitioning variables $U_2$, and $U_1$ or $U_5$. The other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

@@

\begin{enumerate} 
\item Three levels for sample size: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning variable (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of $\approx 0.42$), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of $\approx 0.42$)\footnote{Note that, when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition. However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.}. 
\item Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes. In datasets with treatment-subgroup interactions, the treatment effect difference varied only in nodes 4 and 7, with levels $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$). 
\end{enumerate}

\subsubsection{Variable distributions}
Following the approach of \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$. The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ have the same standard deviation: ${\sigma_U}_k = 10$. Correlations between $U$ variables vary according to the third facet of the data-generating design described above.

@@

\subsubsection{Assessment of performance}

@@

Figure~\ref{fig:xyplot_treesize_interact} indicates that GLMM tree grows trees of about the true tree size in all conditions. On average, tree size increases slightly with sample size for GLMM tree, and dramatically for GLM tree. In the absence of random effects (i.e., $\sigma_{b}=0$), GLM and GLMM trees grow trees of about the same size. However, clear differences in tree size were observed when $\sigma_b > 0$. When sample size is small and random intercepts are not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average. When sample size is larger and random intercepts are not correlated with a partitioning covariate, GLM and GLMM trees are about equally sized. However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially wit larger sample sizes (i.e., 500 or 1000) and when $\sigma_b$ values (i.e., 10). This effect was somewhat stronger when the random intercept was correlated with a covariate that did not appear in the true tree.

@@

To assess the accuracy of trees recovered by both algorithms, we inspected variables and values that were selected for partitioning. For the first split, GLMM tree selected the true partitioning variable ($U_2$) in all datasets, and GLM tree in all but one datasets. The splitting value for $U_2$ selected for the first split was 29.94 for both GLM and GLMM tree, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}).

However, further splits were more accurately recovered by GLMM tree, which accurately recovered the partition in 90.40\% of datasets, and GLM tree in only 61.44\% of datasets. To find the most important predictors of tree accuracy, we fitted a GLM with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors of tree accuracy were algorithm type, sample size, variance of the random intercept, and the level of dependence between the partitioning variables and the random intercept. These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_treeaccuracy}).

Figure~\ref{fig:xyplot_treeaccuracy} indicates that in the absence of random effects, the trees recovered by GLM and GLMM tree were about equally accurate. In the presence of random effects, GLM trees were much less accurate than GLMM trees. This was found for all sample sizes, when random effects were correlated with a partitioning covariate. This effect was somewhat stronger when the correlated $U_k$ was not a true partitioning variable. However, when random intercepts were not correlated with one of the $U_k$ variables, GLMM tree outperformed GLM tree only when sample size was small (i.e., $N = 200$).

@@

\subsubsection{Predictive accuracy of trees}
The predicted treatment-effect differences of GLMM tree were closer to the true differences than those of GLM tree, with a mean correlation of .93 ($\mathrm{SD}=0.12$) for GLMM tree and .88 ($\mathrm{SD}=0.19$) for GLM tree. An ANOVA indicated that the most important predictors of predictive accuracy were algorithm type, sample size, variance of the random intercept, and the treatment-difference effect size. These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_correlations}).

Figure~\ref{fig:xyplot_correlations} shows clear main effects of sample size and treatment-effect differences for both algorithms. In the absence of random effects (i.e., $\sigma_b = 0$), predictions of GLM and GLMM tree were about equally accurate. In the presence of random effects, GLM tree predictions were clearly less accurate than those of GLMM tree when treatment-effect differences were smaller (i.e., Cohen's $d = .5$). This effect was also observed, but much less pronounced, when treatment-effect differences were larger (i.e., Cohen's $d = 1.0$). In other words, GLMM tree outperforms GLM tree in the presence of random effects, especially when sample size and treatment-effect differences are smaller.

@@

\subsubsection{Design}
For assessing Type-I error of GLM and GLMM tree, we generated datasets without treatment-subgroup interactions, in which there is only a main effect of treatment in the population. Put differently, there was only a single global value of $\beta_j = \beta$ in every dataset. For these datasets, we varied the same parameters of the data-generating process as in Study I. However, the sixth facet of the data-generating process had only two levels for these datasets, as there were no `true' partitioning variables in these datasets. Therefore, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2$ = 21,600 datasets without treatment-subgroup interactions were generated.
